---
title: "Nested cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You tuned a model with cross-validation and kept the candidate that scored
best. That score is not an estimate of how the model will do on new data. The
candidate was chosen *because* it scored well on those resamples, so its score
carries the selection along with it, and reporting it overstates what you have.

Nested cross-validation removes the contamination by putting the whole
tune-and-fit procedure inside a second, outer resampling loop. Each outer fold
tunes from scratch on its own analysis data, fits the winner there, and scores
it once on assessment rows that no part of the tuning ever saw. Averaging those
outer scores estimates how the *procedure* — resample, tune, select, fit —
performs on new data.

That last sentence is the whole idea, and it has a consequence worth stating up
front: what comes back is a property of the procedure, never of any one fitted
model. The model you eventually deploy is a separate object, produced further
down this page, and it has no honest performance number of its own. The nested
estimate is what you report in its place — as a statement about the procedure
that produced it, not as that model's score. The section below names exactly
which quantity it is.

```{r libraries, message = FALSE}
library(nestedtune)
library(parsnip)
library(rsample)
library(workflows)
```

```{r deps, include = FALSE}
has_ranger <- requireNamespace("ranger", quietly = TRUE)
```

```{r deps-notice, eval = !has_ranger, echo = FALSE, results = "asis"}
cat(
  "The worked example below fits random forests with the **ranger** package,",
  "which is not installed here, so the rest of this guide has been skipped.",
  "Install ranger and rebuild to see it.\n"
)

knitr::knit_exit()
```

## The design

`nested_resamples()` builds the two-level structure: an outer resampling, with
an inner resampling attached to each outer fold.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

folds
```

Each row is one outer fold. `splits` holds that fold's outer split, and
`inner_resamples` holds an ordinary `rset` built from its analysis rows alone —
which is what the tuning for that fold gets to see.

```{r inner}
folds$inner_resamples[[1]]
```

`mtcars` has `r nrow(mtcars)` rows, which keeps this page fast to build. It is
also, honestly, the situation where nested cross-validation earns its cost:
with little data the tuning step is unstable, and the optimism it introduces is
largest.

## The model and the grid

Anything `tune` can tune, this can tune. Here it is a random forest with two
parameters marked for tuning, and an explicit grid of candidates.

```{r workflow}
rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
grid
```

That is `r nrow(grid)` candidates, each of which will be resampled inside every
outer fold — so the run below fits
`r nrow(folds) * nrow(folds$inner_resamples[[1]]) * nrow(grid)` models for
tuning, plus one per outer fold for scoring. Nested cross-validation is
expensive, and that arithmetic is where the cost lives.

## Running the loop

`nested_tune_grid()` drives the outer loop. For each outer fold it calls
`tune::tune_grid()` on that fold's inner resamples, selects the best candidate,
finalizes the workflow, and fits and scores it on the outer split. Every
statistical step is `tune`'s; what this package contributes is the loop, a
reproducibility contract, and a result that keeps what each fold chose.

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

## What to report, and why

```{r estimate}
est <- collect_metrics(res)
est

rmse_row <- est[est$.metric == "rmse", ]
```

Report that. The RMSE of `r round(rmse_row$mean, 2)`, measured `r rmse_row$n`
times on rows the procedure never touched, estimates one specific quantity, and
it is worth naming it: the **k-fold test error of the tune-and-fit procedure**
(Bayle et al., 2026). That is the error you should expect from resampling,
tuning, selecting and fitting — the whole thing — on a training set the size of
one outer analysis set, and then predicting fresh data drawn like `mtcars`.

Two quantities it is *not*, both easy to mistake it for:

- **Not the risk of the model you deploy.** That model is built further down
  this page on all `r nrow(mtcars)` rows; its own error is a different quantity
  with a different value, and nothing here estimates it. What the nested number
  buys you is the right to describe the procedure that produced that model.
- **Not the same quantity averaged over training sets.** The average — how this
  procedure does across all datasets of this size, rather than on the ones these
  folds happened to draw — is usually what a reader has in mind, and it is much
  the harder thing to get. Luo and Barber (2026) prove that any assumption-free
  test of it needs a dataset many times larger than the training size being
  evaluated; a `r nrow(folds)`-fold outer loop supplies a ratio of
  `r nrow(folds)`/`r nrow(folds) - 1`, which is not that. So this package
  reports the estimate and offers no test of it.

Expect that number to run slightly **pessimistic** — to look a little worse
than the procedure deserves. The reason is arithmetic rather than statistical:
every outer fold trains on its analysis rows only, so every model scored is
built on less data than the model you finally deploy, and less data usually
means worse predictions. Two measurements bracket the size of it. Varma and
Simon (2006) found a nested estimate of 54.2% against a true 50.0% on
deliberately null data with 40 samples — a 4.2-point overshoot they attribute
directly to training on 39 rows rather than 40. Wilimitis and Walsh (2023),
working on 41,121 hospital visits, found nested cross-validation the most
pessimistic of every method they compared, by roughly 1–2% of AUROC and 5–9%
of AUPR against a held-out benchmark. The offset therefore shrinks as folds
get larger, but it does not vanish, and it points the same way in both.

Two cautions on reading that. The gap is a property of the estimator, not a
prediction about any one run: at the sample size on this page it is small next
to fold-to-fold noise, so this particular vignette's numbers could as easily
land the other way, and nothing below is offered as a demonstration of it.
And "pessimistic" is a claim about the procedure's error, never a licence to
adjust the reported figure upward — there is no correction here to apply.

`std_err` is the standard error *of that mean* — the standard deviation of the
`r rmse_row$n` fold scores divided by the square root of how many there were. It
is not the fold-to-fold spread, which is larger by that same factor, and it is
not a confidence interval: this package deliberately ships no inference on the
nested estimate.

That has a consequence for the obvious next thing to do with two of these
numbers. If you run `nested_tune_grid()` on two different workflows and one
comes back lower, nothing in `collect_metrics()` output tells you whether the
difference is real — there is no valid interval here to subtract. Bayle et al.
(2026) give a second, sharper reason not to reach for one: the stability
condition that makes a single procedure's cross-validation interval valid can
fail for the *difference* between two procedures even when it holds for each of
them separately, and it fails hardest when the two are similar — which is
exactly the case where you most want the comparison. Treat a gap between two
nested estimates as a reason to look closer, never as a result.

Now the number **not** to report. Any tuning run — the ones inside these outer
folds, and the one behind the final model further down — carries resampling
scores for every candidate it tried. Those scores are what selection *looked
at*, and the winner's is the best of a set scored on the very resamples that
chose it. It therefore carries an optimistic component of unknown size, and
nothing in the output tells you how large.

What it is not is reliably *worse-looking*. At this sample size the bias is
small next to resampling noise, so a single comparison can land either way — the
structural argument is the reason to distrust the number, not its sign. The
final-model section below puts the two side by side.

Two things the nested estimate does not say, both easy to over-read:

- It is **marginal over selection, not conditional on any one configuration**.
  It describes what happens when you run the tuning procedure, including the
  fact that the procedure sometimes picks differently. It is not a claim about
  the particular `mtry` and `min_n` your deployed model happens to carry.
- It describes **new data drawn like your training data** — not a different
  population, and not the same procedure run at a different sample size.

## What each fold chose

The print method summarized this above, and `.selected` holds it exactly: a list
column of one-row tibbles, one per outer fold, each holding the parameters that
fold's inner tuning chose. Stacked up, with the fold each came from:

```{r selected}
selected <- do.call(rbind, res$.selected)

data.frame(fold = res$id, mtry = selected$mtry, min_n = selected$min_n)
```

```{r instability}
n_mtry <- length(unique(selected$mtry))
n_min_n <- length(unique(selected$min_n))
```

Across `r nrow(res)` outer folds, `mtry` took `r n_mtry` distinct selected
values and `min_n` took `r n_min_n`. Most tools throw this away; nestedtune
keeps it, because it is information about the procedure rather than noise in it.

`autoplot()` draws the same thing, one panel per tuned parameter and one point
per outer fold. A flat row means the folds agreed; scatter means they did not.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

Read it as a statement about how well-determined each tuning choice is at this
sample size. A parameter the folds agree on is one the data picks clearly. A
parameter they split over is one whose value is largely arbitrary — meaning
*whichever* value your final model ends up carrying was not strongly preferred
by the evidence. That is not a defect in the estimate: the outer scores already
average over exactly this variability, which is what makes them describe the
procedure honestly. It is a defect in any story you might tell about the
selected parameters being the right ones.

Disagreement is the expected behaviour, not a symptom, wherever the candidates
in question perform about equally well. Bayle et al. (2026) give the mechanism:
as two candidates' predictions converge, the *difference* between their losses
loses variance faster than the noise in estimating that difference does, so
which of them wins any particular fold becomes close to a coin flip. The run
above shows both faces of it at once — the folds split over `mtry`, whose
candidates the data cannot separate here, and agree unanimously on `min_n`,
where they can.

The per-fold scores are worth a look for the same reason:

```{r per-fold}
per_fold <- collect_metrics(res, summarize = FALSE)
per_fold

fold_rmse <- per_fold$.estimate[per_fold$.metric == "rmse"]
c(sd = sd(fold_rmse), std_err = sd(fold_rmse) / sqrt(length(fold_rmse)))
```

Wide spread across outer folds at this sample size is expected. Note the two
numbers above: `r round(sd(fold_rmse), 2)` is how much the folds actually differ
from each other, and `r round(sd(fold_rmse) / sqrt(length(fold_rmse)), 2)` is
the `std_err` `collect_metrics()` reports — the precision of their *mean*.
Quoting the second as though it described the folds understates their
disagreement.

The other view of `autoplot()` shows that spread, with the dashed line at the
nested estimate — the same number `collect_metrics()` reports, so the figure and
the summary cannot drift apart:

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, with one point per outer fold's score and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```

## The model you deploy

Nothing above produced a model you can predict with, and that is deliberate.
The estimate describes the procedure; the model is a separate object, built by
running that same procedure once more with the whole dataset in hand.

```{r final}
set.seed(3)

final <- nested_final_fit(wf, folds, grid = grid)

final
```

The outer folds play no part here. Their selections are not pooled or voted on
— they belong to the estimate, which describes the procedure across the
instability they reveal.

The trained workflow comes out with `extract_workflow()`, and predicts as any
workflow does:

```{r predict}
predict(extract_workflow(final), new_data = mtcars[1:3, ])
```

Now the comparison promised above. `final$tuning` is the tuning run this model's
parameters were selected from, and its best score is the number a user is most
tempted to report:

```{r comparison}
selection_scores <- collect_metrics(final$tuning)
selection_rmse <- selection_scores[selection_scores$.metric == "rmse", ]
best_selection_rmse <- min(selection_rmse$mean)

data.frame(
  quantity = c("nested estimate (report this)", "best selection-time score"),
  rmse = c(rmse_row$mean, best_selection_rmse)
)
```

The selection-time score is
`r ifelse(best_selection_rmse < rmse_row$mean, "lower", "higher")` than the
nested estimate here — `r round(best_selection_rmse, 2)` against
`r round(rmse_row$mean, 2)`. Do not read the direction as the lesson. With
`r nrow(mtcars)` rows a difference this size is comfortably inside what
resampling noise produces, and the point stands whichever way it falls: the
selection-time number was computed on the very resamples that chose the winner,
so it is not an estimate of performance on anything. It is kept on the object
because it is the record of what selection saw, and `collect_metrics()` will
hand it over without warning you.

The model in hand has no honest number of its own. Everything computable from
its training data was consumed by selecting it or by fitting it.

That is why both objects refuse tune's ranking generics rather than answering
them — for a different reason each. On the loop's results they would rank *outer
folds*, which is not a ranking of anything a user wants. On the final fit there
is only one model and nothing to rank at all; what they would surface is the
selection-time metrics above, dressed as a score.

```{r refusals, error = TRUE}
tune::show_best(res, metric = "rmse")
```

```{r refusals-final, error = TRUE}
tune::select_best(final, metric = "rmse")
```

## Reproducibility

Seed the session before the call, as elsewhere in tidymodels. Neither function
takes a seed of its own:

```{r signatures}
args(nested_tune_grid)
args(nested_final_fit)
```

Each draws and pins its own per-step seeds from the session state instead, and
stores them, so any single piece is reproducible by hand:

```{r seeds}
res$.tuning_seed
final$tuning_seed
```

Because each fold's seeds are fixed by its *position* in the design rather than
by the order the folds happen to run in, the result does not depend on how the
loop is scheduled. And the caller's own random state is put back as it was
found:

```{r rng-restored}
before <- .Random.seed

invisible(nested_final_fit(wf, folds, grid = grid))

identical(before, .Random.seed)
```

`?nested_tune_grid` and `?nested_final_fit` give the exact hand-replication
recipe for each.

## Writing it up

Everything a write-up needs is on the two objects. A minimal, honest report:

```{r writeup, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "> Hyperparameters (`mtry`, `min_n`) were tuned over a %d-point grid by ",
    "%d-fold cross-validation, nested inside a %d-fold outer cross-validation ",
    "of the entire tune-and-fit procedure (n = %d). The outer folds give an ",
    "estimated RMSE of %.2f (SE %.2f) for the procedure. Across those %d ",
    "folds, selection took %d distinct values of `mtry` and %d of `min_n`. ",
    "The deployed model was produced by applying the same procedure to the ",
    "full dataset, which selected mtry = %d and min_n = %d."
  ),
  nrow(grid),
  nrow(folds$inner_resamples[[1]]),
  nrow(folds),
  nrow(mtcars),
  rmse_row$mean,
  rmse_row$std_err,
  nrow(res),
  n_mtry,
  n_min_n,
  final$selected$mtry,
  final$selected$min_n
))
```

The three things that make it honest are the ones this package exists to keep
together: the estimate is attributed to the procedure and not to the model, the
instability is reported rather than hidden, and the deployed model is described
as what it is — the same procedure applied to all the data, carrying no
performance claim of its own.

## References

Bayle, A., Janson, L., & Mackey, L. (2026). The relative instability of model
comparison with cross-validation. *arXiv:2508.04409*.

Luo, Y., & Barber, R. F. (2026). The limits of assumption-free tests for
algorithm performance. *Bernoulli*, 32(3), 2427–2450.

Tibshirani, R. J., & Tibshirani, R. (2009). A bias correction for the minimum
error rate in cross-validation. *The Annals of Applied Statistics*, 3(2),
822–829.

Vabalas, A., Gowen, E., Poliakoff, E., & Casson, A. J. (2019). Machine learning
algorithm validation with a limited sample size. *PLoS ONE*, 14(11), e0224365.

Varma, S., & Simon, R. (2006). Bias in error estimation when using
cross-validation for model selection. *BMC Bioinformatics*, 7, 91.

Wilimitis, D., & Walsh, C. G. (2023). Practical considerations and applied
examples of cross-validation for model development and evaluation in health
care: Tutorial. *JMIR AI*, 2, e49023.
