---
title: "Nested cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You tuned a model with cross-validation and kept the candidate that scored
best. That score is not an estimate of how the model will do on new data. The
candidate was chosen because it scored well on those resamples, so its score
carries the selection along with it, and reporting it overstates what you have.

Nested cross-validation removes the contamination by putting the whole
tune-and-fit procedure inside a second, outer resampling loop. Each outer fold
tunes from scratch on its own analysis data, fits the winner there, and scores
it once on assessment rows that no part of the tuning ever saw. Averaging those
outer scores estimates how the procedure (resample, tune, select, fit) performs
on new data.

What comes back is a property of the procedure, never of any one fitted model.
The model you eventually deploy is a separate object, produced further down
this page, and it has no honest performance number of its own. This page walks
the path from a design to a write-up. What the estimate means, and what it does
not, is the subject of `vignette("estimate")`; `vignette("results")` reads the
whole results object, and `vignette("tuners")` swaps in the other inner
searches.

```{r deps, include = FALSE}
needed <- c("tidymodels", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This guide attaches the **tidymodels** packages and fits random forests",
  "with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this guide has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The design

`nested_resamples()` builds the two-level structure: an outer resampling, with
an inner resampling attached to each outer fold.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

folds
```

Each row is one outer fold. `splits` holds that fold's outer split, and
`inner_resamples` holds an ordinary `rset` built from its analysis rows alone,
which is what the tuning for that fold gets to see.

```{r inner}
folds$inner_resamples[[1]]
```

`mtcars` has `r nrow(mtcars)` rows, which keeps this page fast to build. With
that little data the tuning step is unstable, and the fold-to-fold selections
further down show it directly. This example is not where nesting removes the
most bias; that is wide data searched hard, and `vignette("estimate")` says
why.

## The model and the grid

Anything `tune` can tune, this can tune. Here it is a random forest with two
parameters marked for tuning, and an explicit grid of candidates.

```{r workflow}
rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
grid
```

That is `r nrow(grid)` candidates, each of which will be resampled inside every
outer fold. The run below fits
`r nrow(folds) * nrow(folds$inner_resamples[[1]]) * nrow(grid)` models for
tuning, plus one per outer fold for scoring. Nested cross-validation is
expensive, and that arithmetic is where the cost lives.

## Running the loop

`nested_tune_grid()` drives the outer loop. For each outer fold it calls
`tune::tune_grid()` on that fold's inner resamples, selects the best candidate,
finalizes the workflow, and fits and scores it on the outer split. Every
statistical step is `tune`'s. What this package contributes is the loop, a
reproducibility contract, and a result that keeps what each fold chose.

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

Printing shows the object: one row per outer fold, and the record each fold
left behind. `summary()` says what the run means.

```{r summarize}
summary(res)
```

## What to report

```{r estimate}
est <- collect_metrics(res)
est

rmse_row <- filter(est, .metric == "rmse")
```

Report that. The RMSE of `r round(rmse_row$mean, 2)` is the mean of
`r rmse_row$n` scores, each measured on rows the procedure never touched. It
estimates the error of the whole procedure, resampling, tuning, selecting and
fitting, on data drawn like `mtcars`. It is not the error of the model you
deploy, and `std_err` is the standard error of that mean, not the spread of
the folds and not a confidence interval. `vignette("estimate")` names the
quantity exactly, says why to expect it to run a little pessimistic, and says
why two of these numbers cannot be subtracted to compare workflows.

## What each fold chose

The summary above listed what each fold selected, and `.selected` is where
those choices live: a list column of one-row tibbles, one per outer fold, each
holding the parameters that fold's inner tuning chose. `collect_selections()`
stacks them, with the fold each came from beside it:

```{r selected}
selected <- collect_selections(res)

selected
```

```{r instability}
n_mtry <- n_distinct(selected$mtry)
n_min_n <- n_distinct(selected$min_n)
```

Across `r nrow(res)` outer folds, `mtry` took `r n_mtry` distinct selected
values and `min_n` took `r n_min_n`. Most tools throw this away. nestedtune
keeps it, because it is information about the procedure rather than noise in
it.

`autoplot()` draws those same selections, one panel per tuned parameter and
one point per outer fold. A flat row means the folds agreed. Scatter means
they did not.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

Read it as a statement about how well-determined each tuning choice is at this
sample size. A parameter the folds agree on is one the data picks clearly. A
parameter they split over is one whose value is largely arbitrary, so whichever
value your final model ends up carrying was not strongly preferred by the
evidence. That is expected wherever the candidates perform about equally well,
and `vignette("estimate")` gives the mechanism.

The per-fold scores are worth a look for the same reason:

```{r per-fold}
per_fold <- collect_metrics(res, summarize = FALSE)
per_fold

fold_rmse <- per_fold |>
  filter(.metric == "rmse") |>
  pull(.estimate)

c(sd = sd(fold_rmse), std_err = sd(fold_rmse) / sqrt(length(fold_rmse)))
```

Wide spread across outer folds at this sample size is expected. Note the two
numbers above: `r round(sd(fold_rmse), 2)` is how much the folds actually differ
from each other, and `r round(sd(fold_rmse) / sqrt(length(fold_rmse)), 2)` is
the `std_err` `collect_metrics()` reports, the precision of their mean. Quoting
the second as though it described the folds understates their disagreement.

The other view of `autoplot()` shows that spread, with the dashed line at the
nested estimate. It is the same number `collect_metrics()` reports, so the
figure and the reported estimate cannot drift apart.

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, with one point per outer fold's score and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```

## The model you deploy

Nothing above produced a model you can predict with, and that is deliberate.
The estimate describes the procedure. The model is a separate object, built by
running that same procedure once more with the whole dataset in hand. The
procedure is read from `res` (the design's inner resampling specification, the
grid, the metrics), so you cannot accidentally specify it differently: the
model and the estimate come from one search.

```{r final}
set.seed(3)

final <- nested_final_fit(wf, res)

final
```

The outer folds play no part here. Their selections are not pooled or voted on.
They belong to the estimate, which describes the procedure across the
instability they reveal.

The object predicts directly. `predict()` and `augment()` on it are the trained
workflow's own methods, and `extract_workflow()` returns the workflow itself:

```{r predict}
predict(final, new_data = mtcars[1:3, ])
```

`extract_tune_results()` returns the tuning run this model's parameters were
selected from, an ordinary tune result, and `show_best()` on it gives the
number a user is most tempted to report:

```{r comparison}
best_selection <- extract_tune_results(final) |>
  show_best(metric = "rmse", n = 1)

best_selection

tibble(
  quantity = c("nested estimate (report this)", "best selection-time score"),
  rmse = c(rmse_row$mean, best_selection$mean)
)
```

The selection-time score is
`r ifelse(best_selection$mean < rmse_row$mean, "lower", "higher")` than the
nested estimate here, `r round(best_selection$mean, 2)` against
`r round(rmse_row$mean, 2)`. Do not read the direction as the lesson. With
`r nrow(mtcars)` rows a difference this size is well inside what resampling
noise produces, and the point stands whichever way it falls: the selection-time
number was computed on the very resamples that chose the winner, so it is not
an estimate of performance on anything. It is kept on the object because it is
the record of what selection saw, and tune's readers will hand it over
without warning you.

The model in hand has no honest number of its own. Everything computable from
its training data was consumed by selecting it or by fitting it. That is why
both objects refuse tune's ranking generics rather than answering them. On the
loop's results they would rank outer folds, which is not a ranking of anything
a user wants. On the final fit there is only one model and nothing to rank, and
what they would surface is the selection-time score dressed as a performance
number.

```{r refusals, error = TRUE}
show_best(res, metric = "rmse")
```

```{r refusals-final, error = TRUE}
select_best(final, metric = "rmse")
```

## Where this example sits

Nesting is expensive, and it removes most when the search is large relative to
the data: wide data, a big grid, and preprocessing the loop has to redo. It
removes little when the data is tall and the search is small, and any feature
selection has to sit inside the workflow you hand it or the loop cannot see it.
`vignette("estimate")` gives the measurements behind each of those. By those
lights the example on this page sits at the modest end: `mtcars` has more rows
than columns, the grid has `r nrow(grid)` points, and there is no feature
selection at all. It is here because it builds in seconds, not because it is
where nesting pays best.

## Reproducibility

Seed the session before the call, as elsewhere in tidymodels. Neither function
takes a seed of its own. Each draws and pins its own per-step seeds from the
session state instead, and stores them, so any single piece is reproducible by
hand. The final fit carries the two seeds too, as `tuning_seed` and
`fit_seed`:

```{r seeds}
res$.tuning_seed
final$tuning_seed
```

Because each fold's seeds are fixed by its position in the design rather than
by the order the folds happen to run in, the result does not depend on how the
loop is scheduled. And the caller's own random state is put back as it was
found:

```{r rng-restored}
before <- .Random.seed

invisible(nested_final_fit(wf, res))

identical(before, .Random.seed)
```

`?nested_tune_grid` and `?nested_final_fit` give the exact hand-replication
recipe for each.

## Writing it up

Everything a write-up needs is on the two objects. A minimal, honest report:

```{r writeup, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "> Hyperparameters (`mtry`, `min_n`) were tuned over a %d-point grid by ",
    "%d-fold cross-validation, nested inside a %d-fold outer cross-validation ",
    "of the entire tune-and-fit procedure (n = %d). The outer folds give an ",
    "estimated RMSE of %.2f (SE %.2f) for the procedure. Across those %d ",
    "folds, selection took %d distinct values of `mtry` and %d of `min_n`. ",
    "The deployed model was produced by applying the same procedure to the ",
    "full dataset, which selected mtry = %d and min_n = %d."
  ),
  nrow(grid),
  nrow(folds$inner_resamples[[1]]),
  nrow(folds),
  nrow(mtcars),
  rmse_row$mean,
  rmse_row$std_err,
  nrow(res),
  n_mtry,
  n_min_n,
  final$selected$mtry,
  final$selected$min_n
))
```

The three things that make it honest are the ones this package exists to keep
together: the estimate is attributed to the procedure and not to the model, the
instability is reported rather than hidden, and the deployed model is described
as what it is, the same procedure applied to all the data, carrying no
performance claim of its own.
