---
title: "Nested cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You tuned a model with cross-validation and kept the candidate that scored
best. That score is not an estimate of how the model will do on new data. The
candidate was chosen *because* it scored well on those resamples, so its score
carries the selection along with it, and reporting it overstates what you have.

Nested cross-validation removes the contamination by putting the whole
tune-and-fit procedure inside a second, outer resampling loop. Each outer fold
tunes from scratch on its own analysis data, fits the winner there, and scores
it once on assessment rows that no part of the tuning ever saw. Averaging those
outer scores estimates how the *procedure* — resample, tune, select, fit —
performs on new data.

That last sentence is the whole idea, and it has a consequence worth stating up
front: what comes back is a property of the procedure, never of any one fitted
model. The model you eventually deploy is a separate object, produced further
down this page, and it has no honest performance number of its own. The nested
estimate is what you report for it.

```{r libraries, message = FALSE}
library(nestedtune)
library(parsnip)
library(rsample)
library(workflows)
```

## The design

`nested_resamples()` builds the two-level structure: an outer resampling, with
an inner resampling attached to each outer fold.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

folds
```

Each row is one outer fold. `splits` holds that fold's outer split, and
`inner_resamples` holds an ordinary `rset` built from its analysis rows alone —
which is what the tuning for that fold gets to see.

```{r inner}
folds$inner_resamples[[1]]
```

`mtcars` has `r nrow(mtcars)` rows, which keeps this page fast to build. It is
also, honestly, the situation where nested cross-validation earns its cost:
with little data the tuning step is unstable, and the optimism it introduces is
largest.

## The model and the grid

Anything `tune` can tune, this can tune. Here it is a random forest with two
parameters marked for tuning, and an explicit grid of candidates.

```{r workflow}
rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
grid
```

That is `r nrow(grid)` candidates, each of which will be resampled inside every
outer fold — so the run below fits
`r nrow(folds) * nrow(folds$inner_resamples[[1]]) * nrow(grid)` models for
tuning, plus one per outer fold for scoring. Nested cross-validation is
expensive, and that arithmetic is where the cost lives.

## Running the loop

`nested_tune_grid()` drives the outer loop. For each outer fold it calls
`tune::tune_grid()` on that fold's inner resamples, selects the best candidate,
finalizes the workflow, and fits and scores it on the outer split. Every
statistical step is `tune`'s; what this package contributes is the loop, a
reproducibility contract, and a result that keeps what each fold chose.

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

## What to report, and why

```{r estimate}
est <- collect_metrics(res)
est

rmse_row <- est[est$.metric == "rmse", ]
```

Report that. The RMSE of `r round(rmse_row$mean, 2)` is an approximately
unbiased — if anything slightly conservative — estimate of how this whole
tune-and-fit procedure performs on new data drawn like `mtcars`, measured
`r rmse_row$n` times on rows the procedure never touched. `std_err` is the
spread of those `r rmse_row$n` outer scores, which is a description of the folds
rather than a confidence interval; this package deliberately ships no inference
on the nested estimate.

Now the number **not** to report. Any tuning run — the ones inside these outer
folds, and the one behind the final model further down — carries resampling
scores for every candidate it tried. Those scores are what selection *looked
at*. Each one helped pick the winner, so the winner's score is the smallest of
a set it was chosen from: optimistically biased as a performance claim, by an
amount nothing in the output tells you. The final-model section below puts that
number next to this one.

Two things the nested estimate does not say, both easy to over-read:

- It is **marginal over selection, not conditional on any one configuration**.
  It describes what happens when you run the tuning procedure, including the
  fact that the procedure sometimes picks differently. It is not a claim about
  the particular `mtry` and `min_n` your deployed model happens to carry.
- It describes **new data drawn like your training data** — not a different
  population, and not the same procedure run at a different sample size.

## What each fold chose

The print method summarized this above, and `.selected` holds it exactly: a list
column of one-row tibbles, one per outer fold, each holding the parameters that
fold's inner tuning chose. Stacked up, with the fold each came from:

```{r selected}
selected <- do.call(rbind, res$.selected)

data.frame(fold = res$id, mtry = selected$mtry, min_n = selected$min_n)
```

```{r instability}
n_mtry <- length(unique(selected$mtry))
n_min_n <- length(unique(selected$min_n))
```

Across `r nrow(res)` outer folds, `mtry` took `r n_mtry` distinct selected
values and `min_n` took `r n_min_n`. Most tools throw this away; nestedtune
keeps it, because it is information about the procedure rather than noise in it.

Read it as a statement about how well-determined each tuning choice is at this
sample size. A parameter the folds agree on is one the data picks clearly. A
parameter they split over is one whose value is largely arbitrary — meaning
*whichever* value your final model ends up carrying was not strongly preferred
by the evidence. That is not a defect in the estimate: the outer scores already
average over exactly this variability, which is what makes them describe the
procedure honestly. It is a defect in any story you might tell about the
selected parameters being the right ones.

The per-fold scores are worth a look for the same reason:

```{r per-fold}
collect_metrics(res, summarize = FALSE)
```

Wide spread across outer folds at this sample size is expected, and it is the
same spread `std_err` summarizes above.

## The model you deploy

Nothing above produced a model you can predict with, and that is deliberate.
The estimate describes the procedure; the model is a separate object, built by
running that same procedure once more with the whole dataset in hand.

```{r final}
set.seed(3)

final <- nested_final_fit(wf, folds, grid = grid)

final
```

The outer folds play no part here. Their selections are not pooled or voted on
— they belong to the estimate, which describes the procedure across the
instability they reveal.

The trained workflow comes out with `extract_workflow()`, and predicts as any
workflow does:

```{r predict}
predict(extract_workflow(final), new_data = mtcars[1:3, ])
```

Now the comparison promised above. `final$tuning` is the tuning run this model's
parameters were selected from, and its best score is the number a user is most
tempted to report:

```{r comparison}
selection_scores <- collect_metrics(final$tuning)
selection_rmse <- selection_scores[selection_scores$.metric == "rmse", ]
best_selection_rmse <- min(selection_rmse$mean)

data.frame(
  quantity = c("nested estimate (report this)", "best selection-time score"),
  rmse = c(rmse_row$mean, best_selection_rmse)
)
```

The selection-time score is
`r ifelse(best_selection_rmse < rmse_row$mean, "lower", "higher")` here —
`r round(best_selection_rmse, 2)` against `r round(rmse_row$mean, 2)`. It is
kept on the object because it is the record of what selection saw, not because
it describes this model, and `collect_metrics()` will hand it over without
warning you. The model in hand has no honest number of its own: everything
computable from its training data was consumed by selecting it or by fitting
it.

This is also why `show_best()` and `select_best()` have no method for either
object this package returns. Both would rank outer folds and hand back
something authoritative-looking and meaningless, so they error instead of
answering.

## Reproducibility

There is no `seed` argument anywhere. Seed the session before the call, as
elsewhere in tidymodels, and both functions draw and pin their own per-step
seeds from that state — so a run is identical however the folds are scheduled,
and the caller's random state is left exactly as it was found.

The seeds are stored, so any single piece is reproducible by hand:

```{r seeds}
res$.tuning_seed
final$tuning_seed
```

`?nested_tune_grid` and `?nested_final_fit` give the exact hand-replication
recipe for each.

## Writing it up

Everything a write-up needs is on the two objects. A minimal, honest report:

```{r writeup, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "> Hyperparameters (`mtry`, `min_n`) were tuned over a %d-point grid by ",
    "%d-fold cross-validation, nested inside a %d-fold outer cross-validation ",
    "of the entire tune-and-fit procedure (n = %d). The outer folds give an ",
    "estimated RMSE of %.2f (SE %.2f) for the procedure. Across those %d ",
    "folds, selection took %d distinct values of `mtry` and %d of `min_n`. ",
    "The deployed model was produced by applying the same procedure to the ",
    "full dataset, which selected mtry = %d and min_n = %d."
  ),
  nrow(grid),
  nrow(folds$inner_resamples[[1]]),
  nrow(folds),
  nrow(mtcars),
  rmse_row$mean,
  rmse_row$std_err,
  nrow(res),
  n_mtry,
  n_min_n,
  final$selected$mtry,
  final$selected$min_n
))
```

The three things that make it honest are the ones this package exists to keep
together: the estimate is attributed to the procedure and not to the model, the
instability is reported rather than hidden, and the deployed model is described
as what it is — the same procedure applied to all the data, carrying no
performance claim of its own.
