---
title: "Reading the results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading the results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Every driver in this package returns the same kind of object, a
`nested_results`, and the getting-started guide, `vignette("nested-cv")`,
reads three things off it: the print, the summary and the estimate. This page
reads the whole object. It walks the columns one at a time, runs each of the
readers on one result, shows what a run looks like when one outer fold fails,
says which dplyr verbs keep the class and which shed it, and ends with the two
arguments a censored-regression run adds.

```{r libraries, message = FALSE}
library(nestedtune)
library(parsnip)
library(rsample)
library(workflows)
```

```{r deps, include = FALSE}
has_ranger <- requireNamespace("ranger", quietly = TRUE)
has_recipes <- requireNamespace("recipes", quietly = TRUE)
```

```{r deps-notice, eval = !(has_ranger && has_recipes), echo = FALSE, results = "asis"}
absent <- c("ranger", "recipes")[!c(has_ranger, has_recipes)]

cat(
  "The runs below fit random forests with the **ranger** package and",
  "preprocess with **recipes**, and",
  if (length(absent) == 1L) {
    paste0("**", absent, "** is")
  } else {
    "neither is"
  },
  "installed here, so the rest of this page has been skipped. Install",
  if (length(absent) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

## The run

The design, the workflow and the grid are the guide's, unchanged: five outer
folds of `mtcars`, five inner folds under each, and a random forest with two
parameters marked for tuning.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

## What the columns hold

The object is a tibble with one row per outer fold, and the print above
shows its columns. Each holds one piece of what that fold did.

```{r columns}
names(res)
```

`splits` is the fold's outer split, an ordinary rsample split whose analysis
rows the tuning saw and whose assessment rows scored the fold's model.

```{r splits}
res$splits[[1]]
```

`id` is the fold's label, copied from the design. A repeated design carries
more than one label column, and the object records which columns they are.

```{r id}
res$id
```

`.metrics` is the score of the fold's finalized model on its assessment
rows, one row per metric, in the shape `tune::last_fit()` produces. It is
what `collect_metrics()` averages.

```{r metrics}
res$.metrics[[1]]
```

`.selected` is the candidate the fold's inner tuning chose, a one-row tibble
of the tuned parameters. It is what `summary()`, `agreement()` and the
default `autoplot()` read.

```{r selected}
res$.selected[[1]]
```

`.inner_metrics` is the whole table the inner search scored, every candidate
on every inner resample, averaged. It is what the fold's selection was made
from, and `vignette("tuners")` shows how its shape differs by search.

```{r inner-metrics}
res$.inner_metrics[[1]]
```

`.notes` holds what went wrong inside the fold, in tune's shape with a
`location` naming the stage. A fold that ran cleanly has no rows.

```{r notes}
res$.notes[[1]]
```

`.completed` says whether the fold produced a score. The section on a
failed fold below shows a `FALSE`.

```{r completed}
res$.completed
```

`.tuning_seed` and `.outer_fit_seed` are the two seeds each fold ran under,
drawn at entry and fixed by the fold's position in the design. The guide's
reproducibility section says how to replay one fold by hand from them.

```{r seeds}
res[, c(".tuning_seed", ".outer_fit_seed")]
```

The description of the run itself rides on the object as attributes rather
than columns, and `vignette("tuners")` reads one of them, `procedure`.

## The readers

Printing describes the object in hand. `summary()` says what the run means:
how much of the design ran, what each fold chose, and the estimate.

```{r summary}
summary(res)
```

`collect_metrics()` is the estimate as a table, the mean and standard error
of the outer scores. With `summarize = FALSE` it is the outer scores
themselves, one row per fold and metric, which is `.metrics` stacked with
the fold labels beside it.

```{r collect-metrics}
collect_metrics(res)

collect_metrics(res, summarize = FALSE)
```

`agreement()` counts the selections: one row per distinct combination the
folds chose, with how many completed folds chose it. The most frequent row
describes how stable the tuning procedure's choice was on this data. It is
not the final model's parameters, which come from `nested_final_fit()`
running the procedure once more on the whole dataset.

```{r agreement}
agreement(res)
```

`autoplot()` draws the same two facts. The default view is the selections,
one panel per tuned parameter and one point per outer fold.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, mtry and min_n, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

The other view is the outer scores, one panel per metric and one point per
fold, with a dashed line at the mean that `collect_metrics()` reports.

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, rmse and rsq, with one point per outer fold's score on its assessment rows and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```
