---
title: "Reading the results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading the results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Every driver in this package returns the same kind of object, a
`nested_results`, and the getting-started guide, `vignette("nested-cv")`,
reads three things off it: the print, the summary and the estimate. This page
reads the whole object. It walks the columns one at a time, runs each of the
readers on one result, shows what a run looks like when one outer fold fails,
says which dplyr verbs keep the class and which shed it, and ends with the two
arguments the runs above do not use, `event_level` and `eval_time`.

```{r deps, include = FALSE}
needed <- c("tidymodels", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages and fits random forests",
  "with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this page has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The run

The design, the workflow and the grid are the guide's, unchanged: five outer
folds of `mtcars`, five inner folds under each, and a random forest with two
parameters marked for tuning.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

## What the columns hold

The object is a tibble with one row per outer fold, and the print above
shows its columns. Each holds one piece of what that fold did.

```{r columns}
names(res)
```

`splits` is the fold's outer split, an ordinary rsample split whose analysis
rows the tuning saw and whose assessment rows scored the fold's model.

```{r splits}
res$splits[[1]]
```

`id` is the fold's label, copied from the design. A repeated design carries
more than one label column, and the object records which columns they are.

```{r id}
res$id
```

`.metrics` is the score of the fold's finalized model on its assessment
rows, one row per metric, in the shape `tune::last_fit()` produces. It is
what `collect_metrics()` averages.

```{r metrics}
res$.metrics[[1]]
```

`.selected` is the candidate the fold's inner tuning chose, a one-row tibble
of the tuned parameters. It is what `summary()`, `agreement()` and the
default `autoplot()` read.

```{r selected}
res$.selected[[1]]
```

`.inner_metrics` is the whole table the inner search scored, every candidate
on every inner resample, averaged. It is what the fold's selection was made
from, and `vignette("tuners")` shows how its shape differs by search.

```{r inner-metrics}
res$.inner_metrics[[1]]
```

`.notes` holds what went wrong inside the fold, in tune's shape with a
`location` naming the stage. A fold that ran cleanly has no rows.

```{r notes}
res$.notes[[1]]
```

`.completed` says whether the fold produced a score. The section on a
failed fold below shows a `FALSE`.

```{r completed}
res$.completed
```

`.tuning_seed` and `.outer_fit_seed` are the two seeds each fold ran under,
drawn at entry and fixed by the fold's position in the design. The guide's
reproducibility section shows how to restore them, and the help pages for
`nested_tune_grid()` and `nested_final_fit()` give the recipe for replaying
one fold by hand.

```{r seeds}
res[, c(".tuning_seed", ".outer_fit_seed")]
```

The description of the run itself rides on the object as attributes rather
than columns, and `vignette("tuners")` reads one of them, `procedure`.

## The readers

Printing describes the object in hand. `summary()` says what the run means:
how much of the design ran, what each fold chose, and the estimate.

```{r summary}
summary(res)
```

`collect_metrics()` is the estimate as a table, the mean and standard error
of the outer scores. With `summarize = FALSE` it is the outer scores
themselves, one row per fold and metric, which is each fold's `.metrics`
rows, less `.config`, with the fold label beside them.

```{r collect-metrics}
collect_metrics(res)

collect_metrics(res, summarize = FALSE)
```

`agreement()` counts the selections: one row per distinct combination the
folds chose, with how many completed folds chose it, `n`, and that count as
a share of the completed folds, `prop`. The most frequent row
describes how stable the tuning procedure's choice was on this data. It is
not the final model's parameters, which come from `nested_final_fit()`
running the procedure once more on the whole dataset.

```{r agreement}
agreement(res)
```

`autoplot()` draws the same two facts. The default view is the selections,
one panel per tuned parameter and one point per outer fold.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, mtry and min_n, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

The other view is the outer scores, one panel per metric and one point per
fold, with a dashed line at the mean that `collect_metrics()` reports.

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, rmse and rsq, with one point per outer fold's score on its assessment rows and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```

## When a fold fails

A fold that fails does not end the run. The other folds keep their results,
and the fold that failed is recorded rather than dropped. To show that on the
guide's run, the workflow below adds one preprocessing step a careful reader
might write anyway: a range check on horsepower, which refuses to predict for
a car whose horsepower lies outside the range the model was trained on,
beyond a small slack. One
car, the Maserati Bora, has far more horsepower than any other in `mtcars`,
`r max(mtcars$hp)` against `r sort(mtcars$hp, decreasing = TRUE)[2]` for the
next. The fold that holds it out cannot score it, and that fold fails.

```{r failed-recipe}
rec <- recipe(mpg ~ ., data = mtcars) |>
  check_range(hp)

wf_checked <- workflow(rec, rf)
```

```{r failed-run, message = FALSE}
set.seed(2)

failed <- nested_tune_grid(wf_checked, folds, grid = grid)

stopifnot(sum(!failed$.completed) == 1L)
```

```{r failed-print}
failed
```

The run warns twice as it finishes: tune's own warning from the outer fit
that produced nothing, then this package's, naming the fold. The chunk mutes
tune's progress messages and keeps its warnings. The print counts the
failure, `.completed` is `FALSE` for the fold that failed, and its `.notes`
say where.

```{r failed-notes}
which_failed <- which(!failed$.completed)

failed$.notes[[which_failed]][, c("location", "type", "note")]
```

The first row is this package's own note, and its `location` names the stage:
`r failed$id[which_failed]` failed at the `r failed$.notes[[which_failed]]$location[1]`,
the fit and score on the outer split, after its inner tuning had finished.
The rows after it are tune's notes about the cause, relabelled with the stage
they came from. A fold that fails during inner tuning reads the same way with
`inner tuning` as the stage.

A fold can also complete and carry notes. The Maserati sits in the analysis
set of every other fold, so each of those folds had one inner resample that
held it out, and the range check refused that resample for every candidate.
Some folds carry a second note of the same kind from the low end of the
range, an inner resample that held out a car with less horsepower than any
it trained on. Tuning still returned a candidate, from the inner resamples
that ran, and the fold finished. Its notes say so.

```{r completed-notes}
vapply(failed$.notes, nrow, integer(1))

noted <- which(failed$.completed & vapply(failed$.notes, nrow, integer(1)) > 0)

failed$id[noted[1]]

failed$.notes[[noted[1]]][, c("location", "type", "note")]
```

The label in parentheses names the inner resample, a fold of this outer
fold's own inner design, not another outer fold. So `.completed` being
`TRUE` beside a non-empty `.notes` means exactly this: the fold worked, on
less than the whole inner design it was given.

`summary()` on a run with a failed fold still answers, and warns first. The
chunk below catches the warning to show its class and its message, then
prints the summary it returned.

```{r failed-summary}
partial <- withCallingHandlers(
  summary(failed),
  warning = function(w) {
    cat("class:", class(w)[1], "\n")
    cat(conditionMessage(w), "\n")
    invokeRestart("muffleWarning")
  }
)

partial
```

The warning's class lets a script catch it and decide what a partial run is
worth. The message names how many folds the summary covers and how many the
design asked for, and the summary itself describes the folds that ran: the
selections are those `r sum(failed$.completed)` folds' selections, and the
estimate is the mean over their `r sum(failed$.completed)` scores, with the
failed fold absent rather than filled in. `collect_metrics()` and
`agreement()` warn the same way, and `autoplot()` leaves the failed fold's
place on its axis empty. A run in which no fold completed is different:
`summary()` still describes it, but `collect_metrics()`, `agreement()`,
`autoplot()` and `nested_final_fit()` refuse it, because there is no estimate
to report for a design that did not execute.

## What dplyr keeps and what it sheds

The object is a tibble, so dplyr's verbs work on it, and one rule decides
what they hand back: a verb that only reorders the rows or adds or reorders
columns returns a `nested_results`, and a verb that adds or removes a row, or
drops or overwrites one of the columns above, returns a bare tibble, because
an object that no longer holds what the run produced cannot answer for the
run. Adding a column keeps the class.

```{r dplyr-mutate}
with_rmse <- mutate(
  res,
  rmse = vapply(.metrics, function(m) m$.estimate[m$.metric == "rmse"], 1)
)

class(with_rmse)
```

Dropping a fold sheds it. The natural case is keeping only the folds that
completed on the run above, which removes one row.

```{r dplyr-filter}
completed_only <- filter(failed, .completed)

class(completed_only)
```

So does a column subset that leaves the run's record behind, through base
`[` rather than a dplyr verb, since the same rule governs it.

```{r subset}
class(res[, "id"])
```

Both of those hand back the data and nothing more: a plain tibble print, no
summary of the run, no `collect_metrics()`. That is the point. A four-row table
cannot describe itself as a five-fold design, so it stops describing itself.
To read a partial run, read the run: `collect_metrics(failed)` already
averages the folds that completed, and warns that it did.

## Two more arguments: `event_level` and `eval_time`

Every tuning driver takes two more arguments than the runs above used, and
both reach tune untouched; `nested_final_fit()` reads them back from the
results object instead. `event_level` is for classification: it names which
level of the outcome factor counts as the event, `"first"` unless you say
otherwise, and it sets the same slot on the inner tuning and on the outer
scoring fit at once. A regression or censored-regression run ignores it, as
tune does. `eval_time` is for censored regression, and it changes the shape
of the result, so the rest of this section runs one.

```{r srv-deps, include = FALSE}
has_censored <- requireNamespace("censored", quietly = TRUE)
has_survival <- requireNamespace("survival", quietly = TRUE)
has_srv <- has_censored && has_survival
```

```{r srv-notice, eval = !has_srv, echo = FALSE, results = "asis"}
absent <- c("censored", "survival")[!c(has_censored, has_survival)]

cat(
  "The run below needs the **censored** and **survival** packages, and",
  if (length(absent) == 1L) {
    paste0("**", absent, "** is not")
  } else {
    "neither is"
  },
  "installed here, so this section has been skipped. Install",
  if (length(absent) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)
```

```{r srv-data, eval = has_srv, include = has_srv}
set.seed(51)

n <- 180
x1 <- rnorm(n)
x2 <- rnorm(n)
lp <- 0.9 * x1 - 0.6 * x2
early <- rbinom(n, 1, 0.45)
event_time <- ifelse(
  early == 1,
  runif(n, 0.02, 0.6) * exp(-lp / 6),
  rlnorm(n, meanlog = log(15) - lp, sdlog = 0.8)
)
censor_time <- runif(n, 3, 60)

surv_df <- data.frame(
  time = pmin(event_time, censor_time),
  event = as.numeric(event_time <= censor_time),
  x1 = x1,
  x2 = x2
)
```

```{r srv-data-prose, eval = has_srv, echo = FALSE, results = "asis"}
cat(
  "The data is simulated: two predictors, a time, and an event indicator that",
  "is zero where the observation was censored before its event. The model is a",
  "parametric survival regression whose one tunable is the distribution of the",
  "event times, and the grid names three.\n"
)
```

```{r srv-workflow, eval = has_srv, include = has_srv, message = FALSE}
library(censored)

surv_spec <- survival_reg(dist = tune()) |>
  set_engine("survival") |>
  set_mode("censored regression")

surv_wf <- workflow(survival::Surv(time, event) ~ x1 + x2, surv_spec)

surv_grid <- data.frame(dist = c("weibull", "lognormal", "exponential"))
```

```{r srv-metric-prose, eval = has_srv, echo = FALSE, results = "asis"}
cat(
  "A dynamic survival metric is evaluated at a time: the Brier score at one",
  "time asks how well the model predicts who has survived to that point. So the",
  "metric has one value per time, and `eval_time` is where the times are given.",
  "They reach the inner tuning, which selects on the first of them and warns",
  "once per fold to say so (the chunk mutes those warnings), and the outer",
  "scoring fit, which scores at each.\n"
)
```

```{r srv-run, eval = has_srv, include = has_srv, warning = FALSE}
set.seed(61)

surv_folds <- nested_resamples(
  surv_df,
  outside = vfold_cv(v = 3),
  inside = vfold_cv(v = 3)
)

set.seed(62)

surv_res <- nested_tune_grid(
  surv_wf,
  surv_folds,
  grid = surv_grid,
  metrics = metric_set(brier_survival),
  eval_time = c(0.5, 10)
)

collect_metrics(surv_res)
```

```{r srv-prose, eval = has_srv, echo = FALSE, results = "asis"}
times <- sort(unique(collect_metrics(surv_res)$.eval_time))

cat(sprintf(
  paste0(
    "The estimate carries an `.eval_time` column, and has one row per metric ",
    "and time: the Brier score at %s and at %s, each the mean over the outer ",
    "folds of that fold's score at that time. The per-fold table carries the ",
    "same column, and so does each fold's `.metrics`.\n"
  ),
  times[1],
  times[2]
))
```

```{r srv-per-fold, eval = has_srv, include = has_srv}
collect_metrics(surv_res, summarize = FALSE)

surv_res$.selected[[1]]
```

```{r srv-agreement-prose, eval = has_srv, echo = FALSE, results = "asis"}
cat(
  "Everything else reads as before. `agreement()` counts the selected",
  "distributions the way it counted `mtry` above:\n"
)
```

```{r srv-agreement, eval = has_srv, include = has_srv}
agreement(surv_res)
```

```{r srv-close-prose, eval = has_srv, echo = FALSE, results = "asis"}
cat(
  "`summary()` and the default `autoplot()` read the same `.selected` column,",
  "and `autoplot(type = \"performance\")` draws one panel per metric and",
  "time.\n"
)
```
