---
title: "Choosing the inner tuner"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the inner tuner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The getting-started guide, `vignette("nested-cv")`, tunes each outer fold
over a fixed grid with `nested_tune_grid()`. That is one of five ways the
inner search can run. The other four are tune's Bayesian optimisation and
finetune's two racing methods and its simulated annealing, and each has a
driver here that takes the same design and workflow and returns the same
kind of object. This page runs all four on the guide's example, shows what
each fold records about its search, shows how a tune control object reaches
the inner call, and says what differs from calling tune or finetune
directly.

```{r libraries, message = FALSE}
library(nestedtune)
library(parsnip)
library(rsample)
library(workflows)
```

```{r deps, include = FALSE}
has_ranger <- requireNamespace("ranger", quietly = TRUE)
```

```{r deps-notice, eval = !has_ranger, echo = FALSE, results = "asis"}
cat(
  "The runs below fit random forests with the **ranger** package, which is",
  "not installed here, so the rest of this page has been skipped. Install",
  "ranger and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r deps-optional, include = FALSE}
has_finetune <- requireNamespace("finetune", quietly = TRUE)
has_race <- has_finetune &&
  all(vapply(
    c("lme4", "BradleyTerry2"),
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
```

## The design and the workflow

These are the guide's, unchanged: five outer folds of `mtcars`, five inner
folds under each, and a random forest with two parameters marked for tuning.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

The grid tuner and the racers score the candidates they are given. The
Bayesian and annealing searches propose their own candidates instead, and to
do that they need to know each parameter's range. The default range of
`mtry` is not known until the data is seen, so a search over it fails at
inner tuning in every fold with the message that `mtry` must be a parameter
object without unknowns. The parameter set below fixes that range by hand,
and bounds `min_n` to the grid's range too, because its default runs past
the size of an inner analysis set here.

```{r param-info}
params <- update(
  tune::extract_parameter_set_dials(wf),
  mtry = dials::mtry(c(2L, 8L)),
  min_n = dials::min_n(c(2L, 10L))
)

params
```

## Bayesian optimisation

`nested_tune_bayes()` runs `tune::tune_bayes()` inside each outer fold. The
search starts from a small space-filling set of candidates, scores them on
the inner resamples, and then proposes one new candidate per iteration from a
Gaussian process fitted to the scores so far. `initial` and `iter` are the
sizes of those two stages, kept small here so the page builds quickly.

```{r bayes}
set.seed(2)

bayes <- nested_tune_bayes(
  wf,
  folds,
  param_info = params,
  initial = 4,
  iter = 3
)

bayes
```

Each fold scored `r 4 + 3` candidates on `r nrow(folds$inner_resamples[[1]])`
inner resamples, which is `r nrow(folds$inner_resamples[[1]]) * (4 + 3)` fits
per outer fold for tuning, against the
`r nrow(folds$inner_resamples[[1]]) * nrow(grid)` the guide's grid costs. The
print above notes that the folds did not search the same grid. That is the
nature of the search rather than a fault: each fold proposes candidates from
its own scores, so no two folds need score the same set, and the note is
there so that a reader of the summary does not mistake the selections for a
vote over shared candidates.

`.inner_metrics` holds what the search inside one fold saw. Here is the
first fold's:

```{r bayes-inner}
bayes$.inner_metrics[[1]]
```

```{r bayes-iter, include = FALSE}
bayes_iter <- range(bayes$.inner_metrics[[1]]$.iter)
```

The `.iter` column says which stage each candidate came from: `.iter` of
`r bayes_iter[1]` marks the initial set, and the proposals run up to
`r bayes_iter[2]`. A control that stops the search early, shown further
down, is visible here as a fold whose `.iter` stops short of `iter`.

## Racing

The two racing methods in finetune take the same grid the guide uses and
score every candidate on a few inner resamples first. From then on, after
each further resample, a candidate that is clearly worse than the current
best is dropped, so the fits that a full grid search would spend on losing
candidates are saved. `nested_tune_race_anova()` decides with a repeated
measures ANOVA fitted by `lme4`, and `nested_tune_race_win_loss()` with a
Bradley-Terry model of pairwise wins fitted by `BradleyTerry2`.

```{r race-notice, eval = !has_race, echo = FALSE, results = "asis"}
missing <- c("finetune", "lme4", "BradleyTerry2")
missing <- missing[!vapply(
  missing,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

cat(
  "The racing runs need the **finetune**, **lme4** and **BradleyTerry2**",
  "packages, and",
  if (length(missing) == 1L) "one of them is" else "some of them are",
  "not installed here:",
  paste0(paste0("**", missing, "**", collapse = ", "), "."),
  "The two racing sections",
  if (has_finetune) "have" else "and the simulated annealing section have",
  "been skipped. Install what is missing and rebuild to see them.\n"
)
```

```{r race-anova, eval = has_race}
set.seed(3)

anova <- nested_tune_race_anova(wf, folds, grid = grid)

anova
```

A fold's `.inner_metrics` from a race is the whole grid, and the `n` column
says how many inner resamples each candidate was scored on before it was
dropped or the race ended. The fold that spent the fewest fits is the one
where elimination did the most work:

```{r race-anova-inner, eval = has_race}
fits <- vapply(
  anova$.inner_metrics,
  function(m) sum(m$n[m$.metric == "rmse"]),
  numeric(1)
)
fits

cheapest <- which.min(fits)

anova$.inner_metrics[[cheapest]]
```

```{r race-anova-count, eval = has_race, include = FALSE}
inner_v <- nrow(folds$inner_resamples[[1]])
rmse_rows <- anova$.inner_metrics[[cheapest]]
rmse_rows <- rmse_rows[rmse_rows$.metric == "rmse", ]
eliminated <- sum(rmse_rows$n < inner_v)
```

```{r race-anova-prose, eval = has_race, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "In `%s`, %d of the %d candidates show `n` below the %d inner ",
    "resamples, so they were dropped before the race ended, and the fold ",
    "spent %d fits on tuning where the full grid costs %d.\n"
  ),
  anova$id[cheapest],
  eliminated,
  nrow(rmse_rows),
  inner_v,
  fits[[cheapest]],
  inner_v * nrow(grid)
))
```

The win/loss race records the same table, and reads the same way:

```{r race-win-loss, eval = has_race}
set.seed(4)

win_loss <- nested_tune_race_win_loss(wf, folds, grid = grid)

win_loss$.inner_metrics[[1]]
```

## Simulated annealing

`nested_tune_sim_anneal()` runs `finetune::tune_sim_anneal()` inside each
outer fold. Like the Bayesian search it proposes its own candidates and
needs the parameter set above, but each proposal is a small random move from
the current candidate, accepted when it scores better and sometimes when it
scores worse. finetune prints a log of every move by default; the control
passed below keeps the page quiet, and is the first use here of the `...`
that the next section describes.

```{r anneal, eval = has_finetune}
set.seed(5)

anneal <- nested_tune_sim_anneal(
  wf,
  folds,
  param_info = params,
  initial = 3,
  iter = 3,
  control = finetune::control_sim_anneal(verbose_iter = FALSE)
)

anneal$.inner_metrics[[1]]
```

```{r anneal-iter, eval = has_finetune, include = FALSE}
anneal_iter <- range(anneal$.inner_metrics[[1]]$.iter)
```

```{r anneal-prose, eval = has_finetune, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "`.iter` runs from %d, the initial candidates, to %d. Each fold scored ",
    "%d candidates on %d inner resamples, %d fits for tuning, the same as ",
    "the guide's grid.\n"
  ),
  anneal_iter[1],
  anneal_iter[2],
  3 + 3,
  nrow(folds$inner_resamples[[1]]),
  nrow(folds$inner_resamples[[1]]) * (3 + 3)
))
```

## Passing a control through `...`

None of the drivers has a `control` argument of its own. A control object of
the matching kind, `tune::control_bayes()` for the Bayesian driver,
`finetune::control_race()` and `finetune::control_sim_anneal()` for the
others, is passed as `control` through `...`, and reaches the inner tuning
call in every fold. Here the Bayesian run is repeated under a control that
stops a fold's search after three proposals in a row bring no improvement.

```{r control}
set.seed(6)

stopped <- nested_tune_bayes(
  wf,
  folds,
  param_info = params,
  initial = 4,
  iter = 3,
  control = tune::control_bayes(verbose = FALSE, no_improve = 3)
)

procedure <- attr(stopped, "procedure")
procedure
```

The `procedure` attribute is the record of what ran, and a final fit built
from this result re-runs exactly that. Its `control` element is the control
as it took effect, not as it was passed. Two of its slots are this package's
to set. `allow_par` is `r procedure$control$allow_par` whatever the control
said, because parallelism belongs over the outer folds, and a second pool
inside each fold would contend with the first. `event_level` is set once, by
the driver's own argument: a control left at tune's default takes that
argument's level, and a control naming a level that is neither tune's
default nor the argument's is refused at entry, naming both. One slot is
missing from the record. `seed` is dropped, because the Bayesian search's
seed is the fold's tuning seed, which the fold's own `.tuning_seed` column
already holds, and the driver puts it back on the control at the point the
inner call is made.

## What differs from calling tune or finetune directly

Every statistical step is tune's or finetune's, and the driver adds the loop
around it. The differences a caller meets are these.

A control reaches the inner call through `...`, as above, with `allow_par`
forced off, `event_level` taken from the argument and the Bayesian `seed`
supplied by the fold. Everything else on the control passes through as
given, including the slots that stop a search early, which is why one fold's
`.inner_metrics` can be shorter than another's.

The search's own settings are arguments here rather than control slots.
`iter`, `initial` and `objective` on the Bayesian driver, `iter` and
`initial` on annealing, and `grid` on the racers reach the inner call
unchanged, with one narrowing: `initial` is a count, and an earlier tuning
result, which tune and finetune both accept there, is refused.

What comes back from the inner call is its metrics table and its selection,
never the tuning object itself. `.inner_metrics` is that table, the whole
grid for a race, and `.selected` is the best candidate by the first metric.
So a control's `extract`, `save_pred` and `save_workflow` slots cost their
work inside every fold and return nothing on a nested run. The final fit
keeps its tuning run whole, and that is where those slots' results are
reachable.

The racing drivers refuse at entry when a package their race needs is not
installed, rather than one outer loop's worth of work later when finetune
would ask for it, and they refuse a `burn_in` that no fold's inner design
can meet before any fold runs.

Seeding is the guide's contract on all four: seed the session before the
call, and each fold draws its own tuning and fitting seeds from that state,
so the same seed gives the same result serially and in parallel. The help
page of each driver gives the exact hand-replication recipe for one fold.
