---
title: "Running the outer loop in parallel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the outer loop in parallel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Nested cross-validation fits many models, and the outer folds are independent
of one another. `nested_tune_grid()` and its siblings run those folds on
[mirai](https://mirai.r-lib.org/) daemons when a pool is connected, and
serially otherwise. Nothing in the call changes between the two. This page
starts a pool of two daemons, runs the getting-started guide's loop on it,
shows the result identical to the same run made serially, and then goes
through what the parallel path checks before it dispatches, what it sends to
each daemon, what to do while developing the package, and how a run stops.

`?nested_tune_grid` carries the reference version of each of those under its
parallel section. This page shows them running.

```{r deps, include = FALSE}
needed <- c("tidymodels", "mirai", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages, runs the outer loop on",
  "**mirai** daemons and fits random forests with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this page has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The design and the workflow

These are the guide's, unchanged: five outer folds of `mtcars`, five inner
folds under each, a random forest with two parameters marked for tuning, and
a grid of six candidates. `vignette("nested-cv")` walks through each of them.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

## Starting a pool

Parallelism is switched on by connecting daemons and off by disconnecting
them. There is no argument on the loop functions for it. `mirai::daemons()`
starts the pool, and `mirai::status()` reports how many daemons are
connected. Two or more is the threshold, the same one tune uses.

```{r daemons}
mirai::daemons(2)

mirai::status()$connections
```

Each daemon is a separate R process. It does not inherit the options,
library paths, or environment variables this session set after it was
launched, and it loads nestedtune from an installed library rather than from
this session. Both points matter below.

## The pre-flight

Before any fold is dispatched, the call asks every connected daemon whether it
can load nestedtune, whether its copy defines the same internal functions this
session's copy does, and whether it can load each package the workflow and the
tuner need, here parsnip, workflows and ranger. A daemon that fails any of
those stops the call with a message naming how many daemons are affected and
what is missing, rather than letting the run go ahead and every fold come back
as an opaque worker failure. A daemon that does not answer within the bound is
reported as a non-response, which is a different message, so a slow daemon is
never told to install a package it already has. The bound is thirty seconds by
default and is an R option, `nestedtune.preflight_timeout`, in milliseconds.

That first round trip is also what makes each daemon load the tidymodels
stack, so the first parallel call after starting a pool is the slow one.
Later calls in the same session reuse what the daemons already loaded.

## Running the loop

The call is the guide's call. With a pool connected the outer folds are
dispatched to the daemons, and each fold's inner tuning runs serially on its
daemon, because parallelism inside a fold on top of parallelism across folds
would oversubscribe the cores.

```{r run-parallel}
set.seed(2)

par_res <- nested_tune_grid(wf, folds, grid = grid)

par_res
```

The result carries no mark of how it was computed. That is deliberate,
and the serial run below shows what it buys.

## Stopping the pool

`mirai::daemons(0)` disconnects the daemons. After it the same call runs
serially.

```{r daemons-stop}
mirai::daemons(0)

mirai::status()$connections
```

## The same run, serially

The seed is the same, so the serial run is the same procedure on the same
resamples. Each outer fold draws its tuning seed from the session state at
entry, by its position in the design, before anything is dispatched, and
each daemon then pins the same generator kind the serial path uses. So the
fold's inner tuning sees the same random stream whichever process runs it.

```{r run-serial}
set.seed(2)

ser_res <- nested_tune_grid(wf, folds, grid = grid)
```

```{r identity}
same <- c(
  metrics = identical(par_res$.metrics, ser_res$.metrics),
  selected = identical(par_res$.selected, ser_res$.selected),
  tuning_seed = identical(par_res$.tuning_seed, ser_res$.tuning_seed)
)

same
```

`identical()` on the outer scores, the selections and the tuning seeds
returns `r all(same)` for each: the scores, the selections and the seeds
match, and which run used daemons is not recoverable from either. That is what
lets a pool be a matter of the session rather than of the analysis: a script
gives the same answer on a laptop with no daemons and on a workstation with
many.

## What crosses the wire

A fold's payload is its outer split, its inner resamples and its two seeds;
the workflow, the grid and the tuner's settings travel beside it. Every split
in a design indexes one shared copy of the data, but serializing does not
preserve that sharing, so sending the splits as they are would send one copy
of the data per split. The data is therefore stripped from each split before
dispatch and refilled on the daemon, and what crosses is the fold's row
indices plus one copy of the data per fold. Two things this does not reach,
both of them objects you supply rather than anything the package builds. A
recipe keeps a copy of the data it was created with, and a formula carries the
environment it was written in, so a workflow built inside a function that
holds a large object sends that object with every fold. The recipe's copy is
the recipe's own record and travels with every fold; building the workflow at
the top level avoids the second.

## Developing against daemons

Daemons load nestedtune from an installed library. Under
`devtools::load_all()` the host session runs the source tree, and the daemons
cannot see it. With no installed copy in the library, the pre-flight stops the
call. When the library does hold one, the daemons run that copy, which may be
older than the source the host is running. The pre-flight catches the case
where the older copy is missing a function the host's copy has, and asks you
to reinstall and restart the pool. A running daemon keeps the namespace it
already loaded, so reinstalling underneath it changes nothing until the pool
is restarted.

For development, prime the daemons with the source tree instead:

```{r prime, eval = FALSE}
mirai::everywhere(pkgload::load_all("path/to/nestedtune"))
```

## Interrupting and cancelling

Pressing interrupt at the console while folds are outstanding unwinds the
blocking wait, restores the random-number state, and asks the daemons to stop
the folds they have not finished, so the pool goes idle rather than computing
results nobody will read. Stopping is a request. A fold already inside a
compiled fitting routine may not be interruptible, and one that has nearly
finished may simply finish. Nothing about an interrupted run is recorded,
because a fold that was never given a chance to run has not failed.

Cancelling needs mirai's dispatcher, the process that hands tasks to daemons
and can recall them. `mirai::daemons(n)` starts one by default. A pool started
with `dispatcher = FALSE` has no dispatcher, so the request cannot reach the
daemons and the outstanding folds run to completion. The results of such a
pool are correct, and it is not refused. What it lacks is the ability to
stop, and the call says so at dispatch with a warning of class
`nestedtune_pool_not_cancellable`, once per run.

Here is that pool. The first chunk runs the loop on it with the warning
muted and checks the result against the serial run. The second repeats the
call and catches the warning with `rlang::catch_cnd()` to show its class.

```{r no-dispatcher, warning = FALSE}
mirai::daemons(2, dispatcher = FALSE)

set.seed(2)

nd_res <- nested_tune_grid(wf, folds, grid = grid)

identical(nd_res$.metrics, ser_res$.metrics)
```

```{r no-dispatcher-warning}
set.seed(2)

caught <- rlang::catch_cnd(
  nested_tune_grid(wf, folds, grid = grid),
  classes = "nestedtune_pool_not_cancellable"
)

class(caught)

mirai::daemons(0)
```

The message itself names the remedy: restart the pool with
`mirai::daemons(n)`, which starts a dispatcher by default.

```{r no-dispatcher-message, echo = FALSE, comment = ""}
cat(cli::ansi_strip(conditionMessage(caught)))
```

One case the call cannot tell apart is documented rather than guessed at.
Calling `mirai::daemons(0)` while folds are outstanding produces exactly what
a daemon dying mid-fold produces, so tearing the pool down that way is
recorded as fold failures rather than as a cancellation. The other folds keep
their results.

## What this example cannot show

`mtcars` has `r nrow(mtcars)` rows and the grid has `r nrow(grid)` points,
so each fold takes a fraction of a second, and two daemons loading the
tidymodels stack cost more than the folds save. This page shows the parallel
path working and its result matching the serial one, not a speedup. The
speedup arrives where the folds are expensive: larger data, a bigger grid, a
slower engine, or preprocessing the loop has to redo in every fold. The
arithmetic in `vignette("nested-cv")` says where the cost lives, and what is
parallel here is the outer loop over that cost.
