---
title: "Why nest: a simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why nest: a simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A tuning run reports a score for its winner. That score is the best of many
resampling estimates, each noisy, and picking the best of a noisy set biases
it upward. Nested cross-validation exists to remove that bias: each outer
fold tunes on its own analysis rows and scores the winner on rows the tuning
never saw. `vignette("estimate")` explains the argument. This page shows the
size of the effect on data where the true answer is known.

The design follows Varma and Simon (2006). Draw data with no signal at all,
so that every classifier has the same true accuracy, then tune a learner on
it and compare what the tuning run's best score says against what the nested
estimate says. Repeat over fresh draws, and read the two distributions.

```{r deps, include = FALSE}
has_tidymodels <- requireNamespace("tidymodels", quietly = TRUE)
```

```{r deps-notice, eval = !has_tidymodels, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages, which are not installed",
  "here, so the rest of this page has been skipped. Install tidymodels and",
  "rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## What this page reads

Every replicate fits hundreds of small neural networks, so the page does not
run the simulation. It reads a stored result that the script
`vignettes/articles/why-nest-sim.R` in the package repository produces, and
every number below is read from that object. The script's header says how to
regenerate it.

```{r read}
sim <- readRDS("why-nest.rds")
names(sim)
```

## The design

Each replicate draws `r sim$n` rows of `r sim$p` independent standard normal
features and a label that is a fair coin toss, independent of every feature.
Nothing in the features predicts the label, so any classifier's true accuracy
is `r sim$null_accuracy`, whatever its parameters. Features outnumber rows by
`r round(sim$p / sim$n, 1)` to one, a smaller version of the wide null data
Varma and Simon (2006) simulate.

The learner is a single-layer neural network, `parsnip::mlp()` on the nnet
engine, with `hidden_units` and `penalty` marked for tuning and `epochs` fixed
at `r sim$epochs`. The grid holds `r nrow(sim$grid)` candidates:

```{r grid}
sim$grid
```

The same workflow is tuned twice on each replicate's data.

- The **flat run** is `tune::tune_grid()` over `r sim$v_outer`-fold
  cross-validation of all the rows. Its winner's mean accuracy across the folds
  is the number a tuning run shows first, the figure a reader who stops there
  would report.
- The **nested run** is `nested_tune_grid()` over `r sim$v_outer` outer folds,
  each tuning on `r sim$v_inner` inner folds of its own analysis rows. Its
  estimate is the mean accuracy of each outer fold's winner on that fold's
  assessment rows.

Both see the same grid and the same rows, each run drawing its own fold
assignment. What differs is whether the rows that score the winner took part
in choosing it.

```{r design-table, echo = FALSE}
knitr::kable(
  data.frame(
    setting = c(
      "rows", "features", "candidates", "epochs", "weight cap (MaxNWts)",
      "outer folds", "inner folds", "replicates", "seed"
    ),
    value = c(
      sim$n, sim$p, nrow(sim$grid), sim$epochs, sim$max_nwts, sim$v_outer,
      sim$v_inner, sim$replicates, sim$seed
    )
  ),
  align = "lr"
)
```

## The result

The store holds one row per replicate, with the flat run's best score and the
nested estimate side by side.

```{r results}
head(sim$results)
```

```{r medians}
flat_median <- median(sim$results$flat_best)
nested_median <- median(sim$results$nested_estimate)
c(flat = flat_median, nested = nested_median)
```

Over `r sim$replicates` replicates the median of the flat run's best score is
`r round(flat_median, 3)`, against a true accuracy of `r sim$null_accuracy`.
The median nested estimate is `r round(nested_median, 3)`. The flat number
sits `r round(abs(flat_median - sim$null_accuracy), 3)`
`r if (flat_median < sim$null_accuracy) "below" else "above"` the truth on
data that carries no signal. The nested number sits
`r round(abs(nested_median - sim$null_accuracy), 3)`
`r if (nested_median < sim$null_accuracy) "below" else "above"` it, inside the
tolerance of `r sim$tolerance` the store carries.

```{r figure, fig.width = 7, fig.height = 4, fig.alt = "One panel with two columns of points, one per replicate: the flat run's best accuracy on the left and the nested estimate on the right, each column with a horizontal bar at its median, and a dashed line across the panel at the true accuracy of one half. The left column sits above the line and the right column straddles it."}
long <- bind_rows(
  tibble(
    quantity = "Flat run: best candidate's CV accuracy",
    accuracy = sim$results$flat_best
  ),
  tibble(
    quantity = "Nested estimate",
    accuracy = sim$results$nested_estimate
  )
) |>
  mutate(quantity = factor(quantity, levels = unique(quantity)))

ggplot(long, aes(x = quantity, y = accuracy)) +
  geom_hline(yintercept = sim$null_accuracy, linetype = "dashed") +
  geom_point(
    position = position_jitter(width = 0.15, height = 0, seed = sim$seed),
    alpha = 0.6
  ) +
  stat_summary(
    fun = median, fun.min = median, fun.max = median,
    geom = "crossbar", width = 0.4, linewidth = 0.6
  ) +
  labs(
    x = NULL,
    y = "Accuracy",
    title = "Accuracy on data with no signal",
    subtitle = sprintf(
      "%d replicates; bars at the median; dashed line at the true accuracy of %s",
      sim$replicates, format(sim$null_accuracy)
    )
  ) +
  theme_minimal()
```

```{r counts, include = FALSE}
tally <- function(x) {
  c(
    above = sum(x > sim$null_accuracy),
    at = sum(x == sim$null_accuracy),
    below = sum(x < sim$null_accuracy)
  )
}
flat_tally <- tally(sim$results$flat_best)
nested_tally <- tally(sim$results$nested_estimate)
nested_mean <- mean(sim$results$nested_estimate)
nested_se <- sd(sim$results$nested_estimate) / sqrt(sim$replicates)
```

Each point is one replicate. Both quantities vary from draw to draw, with a
standard deviation of `r round(sd(sim$results$flat_best), 3)` for the flat
run's best score and `r round(sd(sim$results$nested_estimate), 3)` for the
nested estimate. Of the `r sim$replicates` replicates, the flat run's best
score is above the truth on `r flat_tally[["above"]]`, at it on
`r flat_tally[["at"]]` and below it on `r flat_tally[["below"]]`; the nested
estimate is above the truth on `r nested_tally[["above"]]`, at it on
`r nested_tally[["at"]]` and below it on `r nested_tally[["below"]]`.

The nested median sits a little below the truth rather than on it. The
design gives no reason to expect that: every label is an independent coin
toss, so whatever a fold's network predicts, the rows it is scored on agree
with it half the time in expectation, and the nested estimate's expectation
is the null accuracy. The mean nested estimate over the replicates is
`r round(nested_mean, 3)`, `r round(abs(nested_mean - sim$null_accuracy), 3)`
`r if (nested_mean < sim$null_accuracy) "below" else "above"` the truth
against a standard error of `r round(nested_se, 3)`. The median is the
summary the page holds to its tolerance, and on this many replicates it
lands where the draws put it.

## Why the flat number is high

Varma and Simon (2006) give the mechanism. The flat run scores every
candidate on the same folds and reports the maximum. Each of those scores is
an unbiased estimate of an accuracy that, on this data, is the same for every
candidate. The maximum of a set of noisy estimates of one value is not an
unbiased estimate of that value; it is biased upward, and more so the larger
the set. For candidates whose estimates are independent, their expression
for the probability that the minimum error over the grid falls below the
truth is one minus one half to the power of the grid size, so the chance
that it does not is one half to that power, for this page's
`r nrow(sim$grid)` candidates one in
`r format(2^nrow(sim$grid), big.mark = ",", scientific = FALSE)`. Here the
candidates are scored on the same folds, so their estimates are not
independent and that figure overstates the case, but the direction is the
same. The best candidate's score describes how lucky the search was, not how
good the winner is.

The nested run's outer folds are not part of that search. Each fold's winner
is chosen on the inner folds and then scored once on rows the search never
touched, so the score has no maximum taken over it. Its expectation is the
true accuracy of the whole tune-and-fit procedure, which here is the null
accuracy. That is the argument `vignette("estimate")` makes, and the figure
above is what it looks like on data where the truth is known.

## What this page does not show

This is one learner on one design. The size of the gap depends on how many
candidates the search compares, how noisy each candidate's resampling score
is, and how much the candidates' true accuracies actually differ. A grid
whose candidates fit the data very differently, or a learner whose tuning
parameters barely change its fit, would show a different gap, and on data
with real signal the flat number is high for the same reason but the truth
is no longer known. Varma and Simon (2006) report the same design with two
other learners and much wider data, and Ambroise and McLachlan (2002) reach
a matching conclusion for feature selection rather than tuning, from labels
permuted rather than drawn at random.

The nested estimate is not free of every bias. Each outer fold's winner is
fit on fewer rows than the model you would deploy, so on data with signal the
nested number runs slightly pessimistic. `vignette("estimate")` sizes that.
On this page every label is an independent coin toss, so a training set of
any size has the same true accuracy and the nested mean sits within a
standard error of the truth. Varma and Simon (2006), whose null data fixes
twenty rows per class, report a nested estimate 4.2 points pessimistic on
it.

## Provenance

The stored result was produced by `vignettes/articles/why-nest-sim.R` at
commit `r sim$commit` on `r sim$date`, from seed `r sim$seed` under the
`r sim$rng_kind[[1]]` generator. The commit is one on the branch that added
this page, reachable through that pull request's history. On one machine and
one set of package versions, the script writes the same object from the same
seed apart from those two provenance fields; other versions of R or of the
fitting packages may not reproduce the fits. The tolerance the store
carries, `r sim$tolerance`, is the distance from the null accuracy inside
which the nested median is held when the result is regenerated.

## References

Ambroise, C., & McLachlan, G. J. (2002). Selection bias in gene extraction on
the basis of microarray gene-expression data. *Proceedings of the National
Academy of Sciences*, 99(10), 6562–6566.

Varma, S., & Simon, R. (2006). Bias in error estimation when using
cross-validation for model selection. *BMC Bioinformatics*, 7, 91.
