---
title: "Why nest: a simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why nest: a simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A tuning run reports a score for its winner. That score is the best of many
resampling estimates, each noisy, and picking the best of a noisy set biases
it upward. Nested cross-validation exists to remove that bias: each outer
fold tunes on its own analysis rows and scores the winner on rows the tuning
never saw. `vignette("estimate")` explains the argument. This page shows the
size of the effect on data where the true answer is known.

The design follows Varma and Simon (2006). Draw data with no signal at all,
so that every classifier has the same true accuracy, then tune a learner on
it and compare what the tuning run's best score says against what the nested
estimate says. Repeat over fresh draws, and read the two distributions.

## What this page reads

Every replicate fits hundreds of small neural networks, so the page does not
run the simulation. It reads a stored result that the script
`vignettes/articles/why-nest-sim.R` in the package repository produces, and
every number below is read from that object. The script's header says how to
regenerate it.

```{r read}
sim <- readRDS("why-nest.rds")
names(sim)
```

## The design

Each replicate draws `r sim$n` rows of `r sim$p` independent standard normal
features and a label that is a fair coin toss, independent of every feature.
Nothing in the features predicts the label, so any classifier's true accuracy
is `r sim$null_accuracy`, whatever its parameters. Features outnumber rows by
`r round(sim$p / sim$n, 1)` to one, a smaller version of the wide null data
Varma and Simon (2006) simulate.

The learner is a single-layer neural network, `parsnip::mlp()` on the nnet
engine, with `hidden_units` and `penalty` marked for tuning and `epochs` fixed
at `r sim$epochs`. The grid holds `r nrow(sim$grid)` candidates:

```{r grid}
sim$grid
```

The same workflow is tuned twice on each replicate's data.

- The **flat run** is `tune::tune_grid()` over `r sim$v_outer`-fold
  cross-validation of all the rows. Its winner's mean accuracy across the folds
  is the number a tuning run shows first, the figure a reader who stops there
  would report.
- The **nested run** is `nested_tune_grid()` over `r sim$v_outer` outer folds,
  each tuning on `r sim$v_inner` inner folds of its own analysis rows. Its
  estimate is the mean accuracy of each outer fold's winner on that fold's
  assessment rows.

Both see the same grid and the same rows. What differs is whether the rows
that score the winner took part in choosing it.

```{r design-table, echo = FALSE}
knitr::kable(
  data.frame(
    setting = c(
      "rows", "features", "candidates", "epochs", "outer folds",
      "inner folds", "replicates", "seed"
    ),
    value = c(
      sim$n, sim$p, nrow(sim$grid), sim$epochs, sim$v_outer, sim$v_inner,
      sim$replicates, sim$seed
    )
  ),
  align = "lr"
)
```

## The result

The store holds one row per replicate, with the flat run's best score and the
nested estimate side by side.

```{r results}
head(sim$results)
```

```{r medians}
flat_median <- median(sim$results$flat_best)
nested_median <- median(sim$results$nested_estimate)
c(flat = flat_median, nested = nested_median)
```

Over `r sim$replicates` replicates the median of the flat run's best score is
`r round(flat_median, 3)`, against a true accuracy of `r sim$null_accuracy`.
The median nested estimate is `r round(nested_median, 3)`. The flat number
sits `r round(flat_median - sim$null_accuracy, 3)` above the truth on data
that carries no signal. The nested number sits
`r round(abs(nested_median - sim$null_accuracy), 3)`
`r if (nested_median < sim$null_accuracy) "below" else "above"` it, inside the
tolerance of `r sim$tolerance` the store carries.

```{r figure, fig.width = 7, fig.height = 4, fig.alt = "One panel with two columns of points, one per replicate: the flat run's best accuracy on the left and the nested estimate on the right, each column with a horizontal bar at its median, and a dashed line across the panel at the true accuracy of one half. The left column sits above the line and the right column straddles it."}
library(ggplot2)

long <- rbind(
  data.frame(
    quantity = "Flat run: best candidate's CV accuracy",
    accuracy = sim$results$flat_best
  ),
  data.frame(
    quantity = "Nested estimate",
    accuracy = sim$results$nested_estimate
  )
)
long$quantity <- factor(long$quantity, levels = unique(long$quantity))

set.seed(sim$seed)
ggplot(long, aes(x = quantity, y = accuracy)) +
  geom_hline(yintercept = sim$null_accuracy, linetype = "dashed") +
  geom_jitter(width = 0.15, height = 0, alpha = 0.6) +
  stat_summary(
    fun = median, geom = "crossbar", width = 0.4, linewidth = 0.6
  ) +
  labs(
    x = NULL,
    y = "Accuracy",
    title = "Accuracy on data with no signal",
    subtitle = sprintf(
      "%d replicates; bars at the median; dashed line at the true accuracy of %s",
      sim$replicates, format(sim$null_accuracy)
    )
  ) +
  theme_minimal()
```

Each point is one replicate. Both quantities vary from draw to draw, and the
nested estimate varies more (a standard deviation of
`r round(sd(sim$results$nested_estimate), 3)` against
`r round(sd(sim$results$flat_best), 3)`), since each of its outer folds is
scored on `r sim$n / sim$v_outer` rows. The flat run's best score is above
the truth on `r sum(sim$results$flat_best > sim$null_accuracy)` of
`r sim$replicates` replicates; the nested estimate is above it on
`r sum(sim$results$nested_estimate > sim$null_accuracy)`.

The nested median sits a little below the truth rather than on it. That is
what a split of a small sample with no signal does. When the rows a fold
trains on lean toward one label, the network tends to predict that label,
and the rows it is scored on are the rest of the sample, which lean the
other way, so the fold's accuracy falls short of chance. The effect shrinks as the folds grow
and is a property of scoring on held-out rows, not of nesting; it is the
opposite sign from the selection bias the flat number carries.

## Why the flat number is high

Varma and Simon (2006) give the mechanism. The flat run scores every
candidate on the same folds and reports the maximum. Each of those scores is
an unbiased estimate of an accuracy that, on this data, is the same for every
candidate. The maximum of a set of noisy estimates of one value is not an
unbiased estimate of that value; it is biased upward, and more so the larger
the set. Their expression for the probability that the minimum error over a
grid of candidates falls below the truth is one minus one half to the power
of the grid size, which for this page's `r nrow(sim$grid)` candidates is
`r 1 - 0.5^nrow(sim$grid)`, indistinguishable from certainty. The best
candidate's score describes how lucky the search was, not how good the
winner is.

The nested run's outer folds are not part of that search. Each fold's winner
is chosen on the inner folds and then scored once on rows the search never
touched, so the score has no maximum taken over it. Its expectation is the
true accuracy of the whole tune-and-fit procedure, which here is the null
accuracy. That is the argument `vignette("estimate")` makes, and the figure
above is what it looks like on data where the truth is known.

## What this page does not show

This is one learner on one design. The size of the gap depends on how many
candidates the search compares, how noisy each candidate's resampling score
is, and how much the candidates' true accuracies actually differ. A grid
whose candidates fit the data very differently, or a learner whose tuning
parameters barely change its fit, would show a different gap, and on data
with real signal the flat number is high for the same reason but the truth
is no longer known. Varma and Simon (2006) report the same design with two
other learners and much wider data, and Ambroise and McLachlan (2002) reach
the same conclusion from labels permuted rather than drawn at random.

The nested estimate is not free of every bias. Each outer fold's winner is
fit on fewer rows than the model you would deploy, so on data with signal the
nested number runs slightly pessimistic. `vignette("estimate")` sizes that.
On this page there is no signal for a smaller training set to lose, so the
effect has nothing to act on.

## Provenance

The stored result was produced by `vignettes/articles/why-nest-sim.R` at
commit `r sim$commit` on `r sim$date`, from seed `r sim$seed` under the
`r sim$rng_kind[[1]]` generator, and the script writes the same object from
the same seed apart from those two provenance fields. The tolerance the store
carries, `r sim$tolerance`, is the distance from the null accuracy inside
which the nested median is held when the result is regenerated.

## References

Ambroise, C., & McLachlan, G. J. (2002). Selection bias in gene extraction on
the basis of microarray gene-expression data. *Proceedings of the National
Academy of Sciences*, 99(10), 6562–6566.

Varma, S., & Simon, R. (2006). Bias in error estimation when using
cross-validation for model selection. *BMC Bioinformatics*, 7, 91.
